S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light

### 2017-04-05 text/evaluation.tex
move content to text file.

\documentclass[conference]{IEEEtran}

\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage{cases}
\usepackage{listings}
\lstset{language=C++,
    basicstyle=\ttfamily \footnotesize,
    keywordstyle=\color{black}\ttfamily,
    stringstyle=\color{black}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    morecomment=[s][\color{blue}]{/**}{**/},
    showstringspaces=false,
    showtabs=false,
    captionpos=b,
    numbers=left
}

\hyphenation{cluster computational-biology short-read-mapping}

\begin{document}

\title{S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light}

\author{\IEEEauthorblockN{Xiaohui Duan, Kai Xu, Yuandong Chan}
\IEEEauthorblockA{%College of Computer Science and Technology \\
Shandong University\\
Jinan, China\\
Email: \{sunrise.duan, xu\_kai, yuandong.c\}@mail.sdu.edu.cn}
\and
\IEEEauthorblockN{Christian Hundt, Bertil Schmidt}
\IEEEauthorblockA{Johannes Gutenberg University\\
Mainz, Germany\\
Email: \{hundt, bertil.schmidt\}@uni-mainz.de}
\and
\IEEEauthorblockN{Pavan Balaji}
\IEEEauthorblockA{Argonne National Laboratory\\
Argonne, Illinois, USA\\
Email: balaji@anl.gov}
\and
\IEEEauthorblockN{Weiguo Liu}
\IEEEauthorblockA{Shandong University\\
Jinan, China\\
Email: weiguo.liu@sdu.edu.cn\\
}
}

\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
The availability and amount of sequenced genomes have been rapidly growing in recent years due to the adoption of next-generation sequencing (NGS) technologies that allow for high-throughput short-read generation at highly competitive cost. Since this trend is expected to continue in the foreseeable future, the design and implementation of efficient and scalable NGS bioinformatics algorithms are important to research and industrial applications.  In this paper, we introduce S-Aligner---a highly scalable read mapper designed for the \emph{Sunway Taihu Light} supercomputer and its fourth-generation ShenWei many-core architecture (SW26010). S-Aligner employs a combination of optimization techniques to overcome both the memory-bound and the compute-bound bottlenecks in the read mapping algorithm. In order to make full use of the compute power of Sunway Taihu Light, our design employs three levels of parallelism: (1) inter-node parallelism using MPI based on a task-grid pattern, (2) intra-node parallelism using multi-threading and asynchronous data transfer to fully utilize all 260 cores of the SW26010 many-core processor, and (3) vectorization to exploit the available 256-bit SIMD vector registers.  Moreover, we have employed asynchronous access patterns and data sharing strategies during file I/O to overcome bandwidth limitations of the network file system. Our performance evaluation demonstrates that S-Aligner scales almost linearly with approximately 95\% efficiency for up to 13,312 nodes (concurrently harnessing more than 3 million compute cores). Furthermore, our implementation on a single node outperforms the established RazerS3 mapper running on a platform with eight Intel Xeon E7-8860v3 CPUs while achieving highly competitive alignment accuracy.
\end{abstract}

\IEEEpeerreviewmaketitle

\section{Introduction}
\label{Introduction}

{\em Next-generation sequencing} (NGS) technologies have revolutionized the field of computational biology because of their massive throughput at highly competitive cost. Nowadays, NGS data is processed in almost every field of bioinformatics with important applications including personalized cancer treatment and precision medicine.
Modern NGS instruments such as Illumina sequencers \cite{reviewngs} allow for the recording of billions of short DNA fragments per day. The produced \emph{short reads} are usually only a few hundred base pairs (bps) in length compared with complete genomes of typical mammals covering billions of nucleotides. In order to further analyze the set of delocalized reads, they are typically mapped to a given reference genome by computing base-pair level alignments to determine their original positions. This exhaustive probing of possible alignments is a compute-heavy task and thus justifies the need for massively parallel and efficient computation patterns. Moreover, read alignment (or mapping) is crucial for many DNA sequence analysis pipelines such as genotyping, the discovery of single nucleotide polymorphism, or personal genomics.

The na\"ive computation of optimal local alignments between each read and the reference genome is considered prohibitive because of the demanding time complexity of the Smith-Waterman algorithm \cite{sw} that is proportional to the product of the sequences' lengths. Thus, methods based on a {\em seed-and-extend} approach are typically employed to dramatically reduce the number of alignment positions by considering only those regions in the reference genome which contain an exact occurrence of a substring of a read. Among existing read mappers one can find two predominant operation modes: {\em any-best} and {\em all}. Bowtie2 \cite{bowtie2}, BWA \cite{bwa}, GEM \cite{gem}, and CUSHAW \cite{cushaw} run in {\em any-best} mode, which determines only one or a few best alignments (also called {\em intervals}) of each read to the reference genome. Other mappers such as RazerS3 \cite{razers3}, Hobbes2 \cite{hobbes2}, BitMapper \cite{bitmapper}, and mrFAST \cite{mrfast} operate in {\em all} mode, which reports all intervals satisfying a certain constraint, for example, compliance with a predefined edit distance threshold.  Because of their extensive nature all-mappers demand far more computational resources than their any-best counterparts. Nevertheless, when studying gene innovation or phenotypic variation the complete list of alignments is crucial for a thorough analysis.

Consequently, we pursue the design of an all-mapper in order to support a wide range of biomedical applications. Associated runtimes can be dramatically reduced if we target clusters with a large number of compute nodes. In the recent past, the prevalent trend in supercomputing has favored heterogeneous nodes featuring attached coprocessors, such as GPUs or Xeon Phi architectures, because of their vast compute capabilities and low energy footprint. A similar architecture has recently been introduced with the \emph{Sunway Taihu Light} supercomputer, which is based on the neo-heterogeneous SW26010 many-core processor. This cluster consists of 40,960 nodes. Each node provides 260 compute cores organized in four {\em compute groups} (CGs). A total of 10,649,600 cores provides Linpack performance of up to 93 PFlop/s, rendering it the world's fastest supercomputer in the Top500 list of November 2016.

In this paper, we introduce S-Aligner---a highly scalable read mapper targeting the Sunway Taihu Light supercomputer and its SW26010 architecture. To the best of our knowledge, we are the first to investigate the applicability and scalability of read mapping on this architecture. Our approach relies on four major contributions. First, we provide an efficient implementation of Myers' bit-parallel string matching algorithm that harnesses the full parallelization and vectorization potential of the SW26010 many-core processor. Our new implementation is three orders of magnitude faster than a na\"ive single-threaded Myers' version. Second, we exploit fast local device memory via DMA intrinsics for intra-CG communication, which results in a 22$\times$ speedup compared with the non-DMA variant. Third, a highly scalable inter-CG communication scheme is proposed. This results in a parallelization efficiency of over 95\% when using 53,248 CGs. Fourth, we employ asynchronous file loading and data-sharing strategies to effectively hide the latency of the network file system. We further expect that the presented techniques can guide the design and implementation of similar types of applications on the neo-heterogeneous many-core cluster architecture of Sunway Taihu Light.

The rest of this paper is organized as follows: Section~\ref{Related Work} describes required background about the Sunway Taihu Light architecture, related work, and the seed-and-extend approach to read mapping. Details of the design and implementation of S-Aligner are presented in Section~\ref{Implementation}. Performance evaluations in terms of speed, scalability, and alignment quality are carried out in Section~\ref{Evaluation}. Section~\ref{Conclusion} summarizes our conclusions.

\section{Background}
\label{Related Work}

\subsection{Sunway Taihu Light Architecture}
\label{Sunway Taihu Light}

The Sunway Taihu Light has been manufactured by the National Research Center of Parallel Computer Engineering \& Technology (NRCPC) of the People's Republic of China and is located at the Wuxi Supercomputing Center. It provides a theoretical peak performance of 125 PFlop/s and an effective performance-to-power ratio of over 6 GFlop/s per watt. It consists of 40,960 nodes with 1.4 PB attached memory. The interconnection network provides roughly 70 TB/s bisection bandwidth and 12 GB/s point-to-point communication bandwidth.

\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\linewidth]{SW26010}
		\caption{Architecture of the SW26010 processor, which is partitioned into four CGs.}
		\label{fig:SW26010}
	\end{center}
\end{figure}

Each node is equipped with a single SW26010 processor that is subdivided into four {\em compute groups} (CGs). A CG is the basic unit that can be addressed by the job scheduling system. Each CG consists of a single {\em master processor} (MP), 64 {\em slave processors} (SPs), and 8 GB of attached DDR3 shared memory. All nodes can access a shared network file system for loading and storing data. The operating system is a customized Linux flavor running on the MP.  Users may manually launch threads on the SPs in order to parallelize compute-heavy portions of their code. Both the MP and 64 SPs support ShenWei's RISC basic instruction set, which provides scalar and SIMD operations. Moreover, reduction primitives can be called exclusively on the MP, while the SPs exhibit specialized intrinsics for high-precision integer arithmetic. The 8 GB attached shared memory can be accessed by both the MP and the 64 SPs via a memory controller with a shared bandwidth of approximately 136 GB/s. The MP has its own cache (L1 and L2), and each SP can access 64 KB of fast {\em local device memory} (LDM). Data residing in shared memory can be written to LDM using DMA intrinsics and subsequently communicated via a broadcast to the LDM of other SPs. Figure~\ref{fig:SW26010} shows the described hardware layout.

ShenWei's SIMD instruction set features 256-bit wide vector types and a set of corresponding intrinsics provided by the \emph{ShenWei 5 Compiler Collection}.  The MPI library (Sunway MPI) is a derivative of the MPICH implementation, and is compliant with the MPI-3 standard.  Basic threading capabilities for the SPs are provided by the \emph{athread} library.

\subsection{Related Work}
We are not the first to parallelize read mapping on a compute cluster. Among the earliest tools are pBWA \cite{pbwa} and pMap \cite{pmap}. pBWA is an MPI implementation of Version 0.5.9 of the BWA aligner while pMap provides an MPI-based framework for the concurrent execution of popular aligners such as BWA \cite{bwa} and Bowtie \cite{bowtie}. Other implementations exploit MapReduce frameworks, for example, the Hadoop-based SEAL \cite{seal} and BigBWA \cite{bigbwa} tools or the Spark-based SparkBWA suite \cite{sparkBWA}, in order to guarantee fault tolerance. Unfortunately, all these approaches suffer from insufficient scalability when executed on hundreds of nodes. The metaframework parSRA \cite{parSRA} is designed for read mappers written in UPC++.  It combines dynamic scheduling and a virtual file system layer for read distribution to overcome the limitations imposed by pMap and MapReduce-based approaches. merAligner \cite{merAligner} and CUSHAW3-UPC++ \cite{cushaw3upc} are other PGAS-based short-read aligners written in UPC; both demonstrate good scalability. Although pMap and parSRA allow for the execution of single node GPU aligners such as nvBowtie \cite{nvBio}, none of the cited tools is explicitly designed for heterogeneous clusters consisting of thousands of many-core processors. Furthermore, all previous approaches target traditional hardware architectures.

To the best of our knowledge, S-Aligner is the first attempt to implement a fully scalable read mapper specifically designed to fit the characteristics of Sunway Taihu Light. Previous algorithms mapped onto this novel supercomputer have focused mainly on application domains outside bioinformatics, such as Earth system modeling, ocean surface wave modeling, atomistic simulation, and phase-field simulation \cite{sunway}.

\subsection{Seed-and-Extend Approach}
%There are some "all" mappers currently on traditional commercial platforms, RazerS3\cite{razers3} is one of the most famous "all" mapper on x86\_64 CPUs. Hobbes2\cite{hobbes2}, BitMapper \cite{bitmapper} and mrFAST\cite{mrfast} are also read mappers in "all" mode on x86\_64 platforms. BitMapper2 also supports CUDA-enabled GPUs.

Consider a set of reads ${\cal R}$, a reference genome $G$, and an error threshold $e$. The read mapping problem can be defined as follows: Find all substrings $g$ of $G$ that are within edit distance $e$ to some read $R \in {\cal R}$. We call such occurrences $g$ in $G$ {\em matches}.

This problem can be solved by a classical {\em dynamic programming} (DP) approach that calculates the semi-global alignment between each $R \in {\cal R}$ and $G$. Unfortunately, each alignment results in a time complexity proportional to the product of the sequences' lengths, which renders intractable the alignment of a large number of short reads to a reference genome a few billion letters long.  In order to address this problem, most state-of-the art solutions \cite{Reinert} employ the {\em seed-and-extend} approach, which follows a three-stage pipeline.

\begin{description}

\item[Stage 1: Filtration.] This stage identifies promising candidate regions (called {\em seeds}) for each read in $G$. A popular strategy in order to discard large regions of $G$ is based on the fact that if a read $R \in {\cal R}$ is divided into $e+1$ non-overlapping $q$-grams (substrings of length $q = \lfloor \lvert R\rvert\ /(e+1) \rfloor$), then (according to the pigeonhole principle) at least one of them occurs exactly in a match. Such occurrences can be identified quickly by looking them up in precomputed $q$-gram index data structure (also called the {\em reference genome index}), which stores all substrings of length $q$ of $G$.

\item[Stage 2. Verification.] This stage determines whether a seed can actually be extended to a full match within edit distance $e$. This requires the implementation of a {\em verification} algorithm in order to analyze the vicinity of each seed. Most mappers typically apply fast versions of DP-based algorithms for this step.

\item[Stage 3. Alignment.] This stage generates the base-pair level alignment information of a read and its verified intervals in $G$. 
%{\color{red} Note that some mappers just use a traceback of Myers delta matrix and some use Smith-waterman, so we shouldn't mention the algorithm}
\end{description}

Established all-mappers such as RazerS3 \cite{razers3} and mrFAST \cite{mrfast} follow this pipeline. Our profiling of RazerS3 and mrFAST using a typical benchmark data set including the human reference HG19 and 1 million simulated Illumina reads (with a read length of 100 bps) reveals that Stage 2 occupies 67\% and 93\% of the overall runtime, respectively. This shows that a fast implementation of Myers' bit-parallel algorithm for computing the edit distance between a read and an interval is a crucial component when designing efficient all-mappers.

Consider two strings $s$ and $s^\prime$ of length $n$ and $m$, respectively. Their {\em edit distance} is the minimum number of point mutations (i.e. insertions, deletions, or substitutions) required to transform $s$ into $s^\prime$. It can be determined by relaxing the cells of a cost matrix $C$ of size $(n+1) \times (m+1)$ according to the recurrence relation shown in Equation~\ref{update_scheme}, where $0 < i \le n$ , $0 < j \le m$ and the {\em characteristic function} $\chi (x \ne y)$ being 1 if $x \ne y$, and 0 otherwise. Initial conditions are set to $C[i,0] = i$, $C[0,j] = j$, and $C[0,0] = 0$.

\begin{align} \label{update_scheme} 
C[i,j]=\min \begin{cases} C[i-1,j-1]+\chi(s_{i-1} \neq s'_{j-1})\\
C[i-1,j]+1 \\
C[i,j-1]+1
\end{cases}
\end{align}

Myers proposed a bit-vector algorithm \cite{myers} that exploits bit parallelism by encoding the differences (deltas) between adjacent rows and columns in the cost matrix $C$ as defined in Equation~\ref{delta_update}, where $\varDelta h_{i,j}$,  $\varDelta v_{i,j}$, and  $\varDelta d_{i,j}$ are the discrete derivatives of $C[i, j]$ in the horizontal, vertical, and diagonal direction, respectively.

\begin{align} \label{delta_update}
\varDelta v_{i,j}&=C[i,j]-C[i-1,j]   &&\in \{0, \pm1\} \nonumber\\
\varDelta h_{i,j}&=C[i,j]-C[i,j-1]   &&\in \{0, \pm1\} \\
\varDelta d_{i,j}&=C[i,j]-C[i-1,j-1] &&\in \{0, +1\}   \nonumber
\end{align}

Note that the absolute values in Equation~\ref{delta_update} are either 0 or 1. Thus, we can encode the relatively small state space with the help of five vectors using one-hot encoding. Figure~\ref{BitDP} shows an example for the encoding of the vertical derivative $\varDelta v$ and its associated one-hot representations $V^+$ and $V^-$.  With this bit-vector representation the cell updates can be rewritten in terms of logical operations as shown in Equation~\ref{one-hot-myers}:

\begin{alignat}{100} \label{one-hot-myers}
H^-_{i, j} &= \chi(\varDelta h_{i, j} &&= -1)            &&= V^+_{i, j-1} &&\land D^0_{i, j} \nonumber\\
V^-_{i, j} &= \chi(\varDelta v_{i, j} &&= -1)            &&= H^+_{i-1, j} &&\land D^0_{i, j} \nonumber\\
H^+_{i, j} &= \chi(\varDelta h_{i, j} &&= +1)            &&= V^-_{i, j-1} &&\lor \lnot(V^+_{i, j-1} \,\lor D^0_{i, j}) \\
V^+_{i, j} &= \chi(\varDelta v_{i, j} &&= +1)            &&= H^-_{i-1, j} &&\lor \lnot(H^+_{i-1, j}   \lor D^0_{i, j}) \nonumber\\
D^0_{i,j}  &= \chi(\varDelta d_{i, j} &&= \phantom{+}0)  &&= V^-_{i, j-1} &&\lor H^-_{i-1, j} \lor \chi(s_{i-1} = s'_{j-1}) \nonumber 
\end{alignat}

\begin{figure}[!htb]
	\includegraphics[width=1\linewidth]{BitDP}
	\caption{Relations between the original matrix, $\varDelta v$, $V^+$, and $V^-$.  Here (a) shows the original score matrix; light-shaded cells indicate $\varDelta v_{i,j}=+1$ while heavily shaded cells correspond to $\varDelta v_{i,j}=-1$. The vertical derivative $\varDelta v$ shown in (b) is subsequently one-hot encoded by the bit vectors $V^+$ and $V^-$ in (c) and (d), respectively.}
	\label{BitDP}
\end{figure}

\section{Design of S-Aligner}
\label{Implementation}

This section describes the implementation of both the inter-CG and intra-CG  parallelism for the S-Aligner. Also discussed is our bit-level encoding strategy and the exploitation of local device memory.

\subsection{Large-Scale Inter-CG Parallelization}

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=\linewidth]{GridNew}
	\caption{Our task-partitioning and file-loading strategy: (a) overall design; (b) and (c) detailed behavior of processes.}
	\label{TaskGrid}
	%\end{center}
\end{figure}

The highest level of parallelization employs a coarse-grained partitioning scheme over a block distribution of reads and reference genome using MPI. In our experiments we process up to 1.6 TB of read data using up to 13,312 SW26010 nodes executing more than 50,000 dispatched processes. Thus, a na\"{\i}ve partitioning of the read file into 50,000 pieces of approximately 32 MB size is not a suitable solution because of the excessive data replication of the reference genome index. Instead, we employ a partitioning strategy based on a {\em task grid pattern} for the inter-CG parallelization by concurrently assigning pairs of reference genome blocks and read chunks to individual CGs. The two dimensions of the grid are spanned by the reference block identifiers (rows) and read chunk identifiers (columns) as shown in Figure \ref{TaskGrid}(a). Thus, processes in the same row share a unique reference genome (index) block while processes within a column use the same read chunk. Each process executes several cells in one row in order to avoid replication of the relatively big reference index.

\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=0.9\linewidth]{RefDiv}
		\caption{Exemplary partitioning of a reference genome based on concatenation and padded splits.}
		\label{RefDiv}
	\end{center}
\end{figure}

The read input file is partitioned into chunks of fixed size. This can be easily achieved by splitting by lines. For the reference genome, we first concatenate the individual chromosomes to form one long string. We then divide it into blocks of suitable size according to the memory capacity of a CG ($\approx$300 million bps per block). Note that this partitioning scheme is disadvantageous when a chromosome is scattered over two blocks or a read is mapped to the junction point of two chromosomes. This issue can be resolved by padding `N's between chromosomes. Moreover, when dividing a chromosome into two blocks, we store several base pairs in a neighborhood around the break point in both chunks. As a result, a read is always mapped correctly to a contiguous region. Figure~\ref{RefDiv} depicts this approach.

The computing nodes of Sunway Taihu Light are connected to a network file system via a 1 GBit/s interface, while the overall file system bandwidth is $\sim290$ GB/s for the whole cluster. In practice, efficient data distribution patterns have to be employed to reach reasonable performance. The file system bandwidth is low compared with the bisection bandwidth of $\sim70$ TB/s among compute nodes. To address this issue, we employ a {\em group-and-broadcast} data reuse strategy to avoid redundant file system accesses by sharing identical data via Sunway's network among nodes. As tasks are assigned to grid cells, we cluster processes working on the same reference genome block to a row group ({\em ref broadcast group}), while processes working on the same read chunk are clustered to a column group. If a process calculates cells in the first row or first column ({\em read broadcast group}), it loads the corresponding reference block or read chunk from the file system as shown in Figure~\ref{TaskGrid}(b). Subsequently, it broadcasts the loaded data to processes calculating the same row or column, while the other processes wait for the broadcast data as shown in Figure~\ref{TaskGrid}(c). After the data-loading stage, processes can perform their computation concurrently.

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=\linewidth]{AsyncRead}
	\caption{Asynchronous data loading strategy: (a) default case, and (b) case when there is no significantly shorter reference block.}
	\label{AsyncRead}
	%\end{center}
\end{figure}

Our reference genome partitioning strategy illustrated in Figure~\ref{RefDiv} often generates one reference genome block that is significantly shorter than the others. This block is assigned to the first row of our task grid. Thus, processes in the first row take less time during computation and will have sufficient time to load reads for the subsequent round of computation. This approach therefore reduces the idle time of other processes waiting for read data.  Figure \ref{AsyncRead}(a) illustrates this strategy.

We further provide an optional asynchronous data-loading strategy in case that there is no significantly shorter reference block. In this case we add a row of processes that loads only reads. When other processes in the column are computing alignments, these processes exclusively load new read chunks. After loading and computation have been completed, all processes within a column receive the loaded data by means of an \texttt{MPI\textunderscore Bcast}. In our experiments, this approach  reduces the idle time between computing two read chunks by over 90\% when using 8 processes. Figure \ref{AsyncRead}(b) illustrates this strategy.

\subsection{Multithreaded Intra-CG Parallelization}

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=1\linewidth]{FrmWk}
	\caption{Workflow of S-Aligner on a single CG: gray boxes correspond to tasks assigned to SPs while tasks in white boxes are executed on the MP.}
	\label{FrmWk}
	%\end{center}
\end{figure}

The second level of parallelism exploits the threading capabilities of a CG. Our design within a single node is a comprehensive read mapper based on the described seed-and-extend approach using the three-stage pipeline: filtration, verification, and alignment. The filtration step uses a look-up table of $q$-grams of the assigned reference genome block that can be loaded from a preprocessed index file or alternatively calculated on-the-fly by a radix sort-based hash table construction method. After the MP performs the lookup of each non-overlapping $q$-gram of a read, it stores the retrieved intervals in a SIMD-friendly manner allowing for efficient vectorization during the subsequent verification step on the SPs. Verification selects all intervals that comply with the restricted edit distance using Myers' bit-parallel algorithm. Afterwards, we employ a banded version of the Smith-Waterman algorithm on the SPs to align the remaining intervals that have passed verification. Figure~\ref{FrmWk} shows the workflow of our intra-CG implementation.

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=1\linewidth]{AsyncPara}
	\caption{Asynchronous filtration.}
	\label{AsyncPara}
	%\end{center}
\end{figure}

The intra-CG parallelization is realized by a task parallelization scheme implemented using spawn and join calls of the {\em athread} library. The MP executes the filtration stage while the SPs process the verification step. Each read is divided into non-overlapping $q$-grams. The MP looks up those $q$-grams in the reference genome index and returns a number of intervals. The MP copies them to a buffer that is transferred to the LDM of SPs for verification. We employ a dual buffer strategy by allocating two buffers for storing the intervals identified by the MP. When the MP has finished filling one buffer, it waits for the SP thread group to join, and then dispatches an SP thread group to verify these intervals. Subsequently, the MP fills another buffer, while the SP thread group performs verification. These steps are repeated until all reads are processed (see Figure \ref{AsyncPara}). The dual buffer strategy reduces the idle times of SPs and is also used for implementing the subsequent alignment stage.

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=\linewidth]{SPParallel}
	\caption{Framework of our parallelization scheme within one CG: the MP performs filtration and output of alignment results while SPs verify identified candidate locations and generate base-pair level alignments. Gray boxes indicate the intervals verified in the second round of verification (assuming $slice\_num = 4$).}
	\label{SPParallel}
	%\end{center}
\end{figure}

Figure~\ref{SPParallel} illustrates our framework for dividing the computation between the MP and SPs within a single CG. The number of seeds identified for a given read can vary while the verification time of a single seed is constant. Thus, a simple static assignment of seed intervals per read to SPs can cause workload imbalance.  To balance the workload, we introduce the parameter $slice\_num$, which restricts the maximum number of seed intervals to be verified for one read. If there are more than $slice\_num$ seed intervals to be verified for a read, we spawn verification threads with the first $slice\_num$ intervals. The remaining intervals will be verified in a subsequent round of spawns. A value of $slice\_num$ that is too large can reduce the overhead of spawning SP thread groups, but it worsens workload balance, and vice versa. Our default value of $slice\_num = 100$ provides a good trade-off in practice.

%In our experiments, the dual buffer strategy shows a performance improvement between 50\% to 70\% in the verification stage by minimizing the idle times of SPs. This optimization technique is also used for the alignment stage.

%\subsection{Using Intrinsics}

\subsection{SIMD Vectorization}
Implementing Myers' bit-parallel algorithm using ShenWei's SIMD intrinsics requires a bit-level encoding of DNA sequences to efficiently evaluate the characteristic function $\chi(s_{i-1} = s'_{j-1})$ for all $i, j$.

Different approaches have been used in existing aligners: BWA \cite{bwa} employs an {\em array of structures} (AoS) while RazerS3 \cite{razers3} stores a DNA sequence in a profile using one-hot encoding. While the AoS approach is more space-efficient and cache-friendly, the usage of a profile with one-hot encoding supports bit-parallel computation in a more efficient way.  Because of the small size of the LDM and the requirement of bit parallelism, we employ an encoding strategy that combines an AoS with an SoA ({\em structure of arrays}) approach, which we call an AoSoA ({\em array of structures of arrays}). This mixed strategy builds two bit-vectors in an interleaved fashion. Thus, we can fetch the encoded sequence by one DMA-intrinsic and calculate $\chi(s_{i-1} = s'_{j-1})$ in terms of two logical equality operations. Figure~\ref{MixPack}  illustrates the encoding strategy.

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=0.9\linewidth]{MixPack}
	\caption{Bit-level encoding strategies for DNA sequences. For simplicity, we use a 4-bit word instead of 64-bit word for illustrating our combined AoSoA strategy (bottom).}
	\label{MixPack}
	%\end{center}
\end{figure}

When updating the five bit-vectors according to Equation~\ref{one-hot-myers} along the columns in a bit-parallel fashion, a circular dependency needs to be resolved: $D^0_{i,j}$ depends on $H^-_{i-1, j}$, which in turn depends on $D^0_{i-1,j}$ (a value we have not computed yet). Myers \cite{myers} has shown that this can be solved by using logical operations and an addition. ShenWei's SIMD instructions support 256-bit integer arithmetic, such as \texttt{simd\_uaddo\_take\_carry} (short for SIMD unsigned octa-word add, taking carry) which adds two 256-bit operands and returns a 256-bit unsigned integer. Thus, different from implementations of Myers' algorithm on other architectures (e.g., \cite{chacon}), our implementation does not require additional instructions to process carries generated within a SIMD lane. Unfortunately, there is no 256-bit compare instruction. Thus, we have implemented comparisons in terms of shifting operations.

Furthermore, we have simplified the core loop as much as possible in order to avoid branching statements. This action results in an implementation consisting of 23 intrinsics as shown in Figure \ref{cores}, where \texttt{resi32} is set to $|read|\mod 32 - 1$.

\begin{figure}[!htb]
    \begin{lstlisting}[frame=single, xleftmargin=3.5ex]
t1    = simd_vxorw(ref_hi,read_hi[k]);
t2    = simd_vxorw(ref_lo,read_lo[k]);
X     = simd_vbisw(t1,t2); 
X     = simd_vxorw(X,one); 
X     = simd_vbisw(X,VP[k]);
D0    = simd_vandw(X,VP[k]);
D0    = simd_uaddo_take_carry(D0,VP[k]);
D0    = simd_vandw(D0,VP[k]);
D0    = simd_vandw(D0,X);
HN    = simd_vandw(VP[k],D0);
HP    = simd_vbisw(VP[k],D0);
HP    = simd_vxorw(HP,one);
HP    = simd_vbisw(HP,VN[k]);	
X     = simd_sllow(HP,resi32);
X     = simd_vbisw(X,pr_HP);
pr_HP = simd_srlow(HP,255);
VN[k] = simd_vandw(X,D0);
VP[k] = simd_vbisw(X,D0);
VP[k] = simd_vxorw(VP[k],one);
t2    = simd_sllow(HN,resi32);
t2    = simd_vbisw(t2,pr_HN);
pr_HN = simd_srlow(HN,255);
VP[k] = simd_vbisw(t2,VP[k]);	
    \end{lstlisting}
%	%\begin{center}
%	%{\scriptsize
%	{\footnotesize
%		\begin{tabular}{| l |}\hline
%			\begin{minipage}{0.95\hsize}
%				\verb| |\\
%				\verb|  |~~1. 	t1 = simd\_vxorw(ref\_hi,read\_ hi[k]);\\
%				\verb|  |~~2.	t2 = simd\_vxorw(ref\_lo,read\_ lo[k]);\\
%				\verb|  |~~3. 	X = simd\_vbisw(t1,t2); \\
%				\verb|  |~~4.   X = simd\_vxorw(X,one); \\
%				\verb|  |~~5.  	X = simd\_vbisw(X,VP[k]); \\
%				\verb|  |~~6.  	D0 = simd\_vandw(X,VP[k]); \\
%				\verb|  |~~7.   D0 = simd\_uaddo\_take\_carry(D0,VP[k]); \\
%				\verb|  |~~8.   D0 = simd\_vandw(D0,VP[k]);\\
%				\verb|  |~~9.  	D0 = simd\_vandw(D0,X);\\
%				\verb|  |~10. 	HN = simd\_vandw(VP[k],D0); \\
%				\verb|  |~11.   HP = simd\_vbisw(VP[k],D0); \\
%				\verb|  |~12.  	HP = simd\_vxorw(HP,one);\\
%				\verb|  |~13.   HP = simd\_vbisw(HP,VN[k]);\\	
%				\verb|  |~14.	X = simd\_sllow(HP,resi32);\\
%				\verb|  |~15.   X = simd\_vbisw(X,pre\_HP);\\
%				\verb|  |~16.	pre\_HP = simd\_srlow(HP,255); \\
%				\verb|  |~17.	VN[k] = simd\_vandw(X,D0);\\
%				\verb|  |~18.	VP[k] = simd\_vbisw(X,D0); \\
%				\verb|  |~19.	VP[k] = simd\_vxorw(VP[k],one);\\
%				\verb|  |~20.	t2 = simd\_sllow(HN,resi32);\\
%				\verb|  |~21.	t2 = simd\_vbisw(t2,pre\_HN); \\
%				\verb|  |~22.	pre\_HN = simd\_srlow(HN,255);\\
%				\verb|  |~23.	VP[k] = simd\_vbisw( t2 ,VP[k]);\\	
%			\end{minipage}\\
%			\hline
%		\end{tabular}
%	}
%	%\end{center}
	\caption{SIMD core instructions used to implement Myers' algorithm on ShenWei. The reference sequence is presented as \texttt{ref\_hi}, the high-bit of the pattern, and \texttt{ref\_lo}, the low-bit of the pattern. The read sequence is stored in the vectors \texttt{read\_hi[k]} and \texttt{read\_lo[k]}. The variables \texttt{D0}, \texttt{HN}, \texttt{HP}, \texttt{VN}, \texttt{VP} refer to the corresponding variables in Equation~\ref{one-hot-myers}.}
	\label{cores}
\end{figure}

\subsection{Exploiting Local Device Memory}

\begin{figure}[!htb]
	%\begin{center}
	\includegraphics[width=1\linewidth]{AsyncTrans}
	\caption{Asynchronous data transfer to LDM.}
	\label{AsyncTrans}
	%\end{center}
\end{figure}

Since SPs do not have any cache and the latency to access the DDR3 shared memory is high, the usage of the explicitly managed LDM is crucial.  DMA fetching is the most efficient way to transfer data between main memory and LDM (i.e., significantly faster than using functions such as \texttt{memcpy}). DMA calls are handled by the memory controller, and SPs can continue to perform computation. Thus, we can overlap data transfers from shared memory to LDM and the verification of intervals using Myers' algorithm by using asynchronous DMA-fetching intrinsics presented by ShenWei.

Figure \ref{AsyncTrans} shows our framework for asynchronous data transfer from DDR3 shared main memory to the LDM of SPs. We allocate two buffers in LDM. When an interval in one buffer is verified, the subsequent interval is being fetched to the other buffer using DMA intrinsics. The SP busy-waits for the completion of DMA-fetching before it starts the next verification.

Here we use a busy-waiting strategy because the time used for DMA fetching is usually shorter than the time required for verification. Thus, in the common case the DMA reply word needs to be checked just once in order to pass the busy-wait loop. Our experimental results show that our asynchronous data transfer implementation can almost hide the DMA-fetching latency completely and gains a performance improvement of a factor of 22 compared with an implementation based on \texttt{memcpy}.

\section{Performance Evaluation}
\label{Evaluation}
The performance of S-Aligner has been evaluated on the Sunway Taihu Light supercomputer. We use GRCh38\footnote{available at http://hgdownload.cse.ucsc.\\edu/downloads.html} as the human reference genome.  For the read input data sets we use either ERR013135\footnote{available at ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR013/\\ERR013135} or reads simulated by Mason \cite{mason} or wgsim \cite{wgsim}.\footnote{Note that Mason usually generates reads of higher quality but does not support long reads.}  We first evaluate the performance in terms of runtime and mapping accuracy on a single node. Subsequently, we evaluate the scalability of our implementation by varying the number of utilized nodes up to 13,312.

>>>>

\section{Performance Evaluation}
\label{Evaluation}

The performance of S-Aligner has been evaluated on the Sunway Taihu
Light supercomputer. We use GRCh38\footnote{available at
  http://hgdownload.cse.ucsc.\\edu/downloads.html} as the human
reference genome.  For the read input data sets we use either
ERR013135\footnote{available at
  ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR013/\\ERR013135} or reads
simulated by Mason \cite{mason} or wgsim \cite{wgsim}.\footnote{Note
  that Mason usually generates reads of higher quality but does not
  support long reads.}  We first evaluate the performance in terms of
runtime and mapping accuracy on a single node. Subsequently, we
evaluate the scalability of our implementation by varying the number
of utilized nodes up to 13,312.
